A Simple and Useful Model of Virtual Memory Performance

### 2010-07-01 article.tex
It's beginning to look like a paper...

\documentclass[11pt]{article}
\usepackage[letterpaper]{geometry}
\usepackage{url}

\title{A Theoretical Approach to Virtual Memory}
\author{Peter Boothe\\
Manhattan College\\
\url{peter.boothe@manhattan.edu}}
\date{Submission to SODA 2011} % or maybe ALENEX2011

\newcommand{\heapsort}{{\sc HeapSort}}
\newcommand{\quicksort}{{\sc QuickSort}}
\newcommand{\bubblesort}{{\sc BubbleSort}}

@@

As memory requirements begin to outstrip available memory on a machine, the
operating system will allocate virtual memory to supplement the available
physical memory.  We present, argue for, and experimentally validate a simple
analysis technique which allows algorithms researchers to analyze the behavior
of an algorithm when using virtual memory.  In particular, the penalty for
fetching a page from virtual memory can be extreme, and can make
previously-small constants quite large. Thus, it would be of great use to be
able to easily estimate, for a given algorithm, the number of times the
algorithm will have to pause to retrieve virtual memory off the hard drive.

@@

In the worst case, every memory access will have to go to the hard drive.
This, however, is not an informative worst-case analysis.  The worst-case page
fault predictions do not reflect the real-world behavior of virtual memory ---
they reflect a situation in which every cache access is a cache miss!  Instead,
our method is to simply count the number of non-adjacent memory and array
accesses.  Using this, we find that we also end up providing a quantitative
measure of ``cache friendliness'', which is a term often used, but seldomly
rigorously defined.  We will demonstrate our method using three classic sorting
techniques: \heapsort, \quicksort, and \bubblesort.

Our analysis of \heapsort is the easiest.  The \heapsort algorithm is simply 
\begin{verbatim}
heapsort(array, size)
    make-heap(array, size)
    for i = size-1 ... 0
        array[i] = remove-min(array, i+1)
    return array
\end{verbatim}

>>>>

\documentclass[11pt]{article}
\usepackage[letterpaper]{geometry}
\usepackage{url}
\usepackage{salgorithm}

\title{A Simple and Useful Model of Virtual Memory Performance}
\author{Peter Boothe and Gregory Rae\\
Manhattan College\\
\url{peter.boothe@manhattan.edu}
\\
\url{greg.rae@gmail.com}
}
\date{Submission to SODA 2011} % or maybe ALENEX2011

\newcommand{\heapsort}{{\sc HeapSort}}
\newcommand{\quicksort}{{\sc QuickSort}}
\newcommand{\mergesort}{{\sc MergeSort}}
\newcommand{\NAA}{\textrm{NAA}}

@@

As memory requirements begin to outstrip available memory on a machine, the
operating system will allocate virtual memory to supplement the available
physical memory.  We present, argue for, and experimentally validate a simple
model and analysis technique which allows algorithm developers to analyze the
behavior of an algorithm when virtual memory is used.  In particular, the
penalty for a page fault in virtual memory can be extreme, and can make
previously-small constants quite large. Thus, it would be of great use to be
able to easily estimate, for a given algorithm, the number of times the
algorithm will have to wait for the operating system to retrieve virtual memory
pages off the hard drive.

@@

In the worst case, every memory access will have to go to the hard drive.
This, however, is not an informative worst-case analysis.  The worst-case page
fault predictions do not reflect the real-world behavior of virtual memory ---
they reflect a situation in which every cache access is a cache miss! Our
model predicts that all non-adjacent memory accesses have a uniform non-zero probability of causing a page fault.

Our method of estimating the number of page faults is therefore to count the
number of non-adjacent memory accesses.  Using this, we find that we also end
up providing a quantitative measure of ``cache friendliness'', which is a term
often used, but seldom rigorously defined.  We demonstrate our method
using three classic sorting techniques: \heapsort, \quicksort, and \mergesort.

\subsection{\heapsort\ Page Fault Analysis}
Our analysis of \heapsort\ is the easiest.  The \heapsort\ algorithm, which we
take from Cormen {\it et al.}\cite{clrs}, is defined, with all of its
supporting functions, in Figure~\ref{fig:hsort}.

\begin{figure}
\begin{algorithm}
\Algorithm{HeapSort}$(array, size)$\+\\
    Build-Max-Heap$(array, size)$\\
    \For $i \gets size$ \DownTo\ $2$\+\\
        exchange $array[1] \leftrightarrow array[i]$\\
        $size \gets size - 1$\\
        Max-Heapify$(array, 1, size)$\-\\
    \Return $array$\-\\
\\
\Algorithm{Build-Max-Heap}$(array, size)$\+\\
    \For $i \gets \lfloor \frac{size}{2} \rfloor$ \DownTo\ $1$\+\\
        Max-Heapify$(array, i, size)$\-\-\\
\\
\Algorithm{Max-Heapify}$(array, i, size)$\+\\
    $l \gets 2*i$\\
    $r \gets 2*i + 1$\\
    $largest \gets i$\\
    \If $l < size$ and $array[l] > array[largest]$\+\\
            $largest \gets l$\-\\
    \If $r < size$ and $array[r] > array[largest]$\+\\
            $largest \gets r$\-\\
    \If $largest \neq i$\+\\
        exchange $array[i] \leftrightarrow array[largest]$\\
        Max-Heapify$(array, largest, size)$
\end{algorithm}

\caption{\heapsort\ from Cormen {\it et al.}\cite{clrs}.}
\label{fig:hsort}
\end{figure}

In the pseudocode of \heapsort, we note that, with the exception of the very
top of the heap, at no point do we ever access memory adjacent to the most
recently used memory location.  Therefore, we predict that the number of page
faults will be proportional to the number of memory accesses, which in the case
of \heapsort\ is proportional to its runtime of $O(n \log n)$.

\subsection{\quicksort\ Page Fault Analysis}
The \quicksort\ algorithm will be dealt with later.

\subsection{\mergesort\ Page Fault Analysis}
The \mergesort\ algorithm, which we also
take from Cormen {\it et al.}\cite{clrs}, is defined, with all of its
supporting functions, in Figure~\ref{fig:msort}.  In our analysis, the only consideration is the number and location of memory accesses, and memory accesses only occur in the ``merge'' portion of the algorithm.

\begin{figure}
\begin{algorithm}
\Algorithm{MergeSort}$(array, p, r)$\+\\
    \If $p < r$\+\\
        $q \gets \lfloor\frac{p+r}{2}\rfloor$\\
        MergeSort$(array, p, q)$\\
        MergeSort$(array, q+1, r)$\\
        Merge$(array, p, q, r)$\-\-\\
\\
\Algorithm{Merge}$(array, p, q, r)$\+\\
    $n_1 \gets q - p + 1$\\
    $n_2 \gets r - q$\\
    create arrays $L[1\ldots n_1+1]$ and $R[1\ldots n_2+1]$\\
    \For $i \gets 1 \ldots n_1$\+\\
        $L[i] \gets array[p+i-1]$\-\\
    \For $j \gets 1 \ldots n_2$\+\\
        $R[j] \gets array[q+j]$\-\\
    $L[n_1+1] \gets \infty$, $R[n_2+1] \gets \infty$\\
    $i \gets 1$, $j \gets 1$\\
    \For $k \gets p \ldots r$\+\\
        \If $L[i] \le R[j]$\+\\
            $array[k] \gets L[i]$\\
            $i \gets i+1$\-\\
        \Else\+\\
            $array[k] \gets R[j]$\\
            $j \gets j+1$
\end{algorithm}

\caption{\mergesort\ from Cormen {\it et al.}\cite{clrs}.}
\label{fig:msort}
\end{figure}

In merging, the first step is to access the beginning and middle of the
passed-in array (two non-adjacent accesses). Then, two extra arrays are
created, the contents of the passed-in array are copied over to the new arrays
(many accesses, only the first access to each array is non-adjacent).  Then, in
the merge step, the contents of the two new arrays are merged back into the
main array (many accesses, but only the first access to each array is
non-adjacent).

In total, each call to the merge function will require five non-adjacent memory
accesses.  Therefore, over the whole of the algorithm, we find that the number
of non-adjacent accesses ($\NAA(n)$) is $$\NAA(n) = 5 + 2  \NAA(\frac{n}{2})$$
Immediately, by the master theorem, we can see that $\NAA(n) \in O(n)$, and
that we therefore expect the number of page faults to grow linearly with the
size of the input to \mergesort.

\section{Experimental Validation of Our Model}
\subsection{Experimental Setup}
\subsection{\heapsort}
\subsection{\quicksort}
\subsection{\mergesort}

\section{Conclusion}

\bibliographystyle{abbrv}
\bibliography{bibliography}
